Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-13 rebuttal.tex

\fbox{\textbf{Reviewer 2}}

\comment{The authors mention limitations of SMILES implementations,
  but do not mention canonicalization in this context, often using the
  Morgan algorithm published in 1965 in J. Chem. Doc.}
\response{We have updated the text on page 3 to note the Morgan
  algorithm in the context of chemical structure representation}

\comment{The authors state that 3D information is lost when only
  considering the molecular graph. However, they do not make it clear
  that the 3D space is implicitly encoded in the molecular graph –
  this is essential to cover properly.}
\response{We have update the text on page 2 to note that 3D
  information is implicit in 2D representations}

\comment{ I entirely dispute that cheminformatics methods have been
  'closely guarded secrets of companies...' since many methods have been
  published in the public domain. The authors suggest that only in the
  last decade has the field gained access to freely available
  software, etc.}
\response{We agree that secret was probably not the best choice of
  words. However we do believe that much cheminformatics software and
  data was proprietary till recently. We have updated the text on
  pages 7 and 8 to expand on this and provide a more detailed
  discussion}

@@

\comment{P7, c2, l51: why is chemically in inverted commas and not
  algorithmically? I would prefer inverted commas for neither.}
\response{Quotes have been removed}

\comment{P8, c1, l10: ELN's should be ELNs.}
\response{Corrected}

\comment{P8, c1, l5: the users emphasise open source, open data but
  this should be an article on all of cheminformatics not just a
  particular subset.}
\response{We have restructured the section on
  open source and open data. It is true that we still focus on open
  source/open data, because this ensures that a reader will be able to
  explore cheminformatics problems with a minimum of encumbrances}

\comment{P7, c1, l9: the authors make an interesting point here
  regarding why cheminformatics software has historically been from
  commercial suppliers. Given the scope of this manuscript it would be
  interesting to expand on this discussion.} 
\response{We have
  restructured the Conclusions section to expand on the role of
  cheminformatics in industry vis a vis bioinformatics, and why much
  of cheminformatics tools and data have been commercial, in contrast
  to freely available tools and data in bioinformatics}

### 2011-11-13 paper.tex
restructured conclusions. responded to reviewer 2

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of the development of cheminformatics from the 1960s until
the new millenium was performed in secret, because of competition in
the pharmaceutical industry.  While chemists were among the first
scientists to embrace web-based tools of knowledge management,
research into fundamental algorithmic questions of structure
enumeration and database management (and a host of other topics) have
suffered, because of limited academic access to either tools or data.
As we enter the second decade of the 21st century, however, this
landscape is rapidly shifting.  There now exist free-of-charge
open-source cheminformatics toolkits, and open databases with tens of
millions of compounds.  We believe the prospects for academic and
industrial collaboration between chemists and computer scientists are
bright, and we would like to encourage more computer scientists to
participate in cheminformatics research.  There is an admittedly steep
learning curve, as the structural issues faced by chemists are often
too detailed and subtle to admit as clean an abstraction as
``algorithms on a string,'' the way many bioinformatics algorithms can
be abstracted into a computer science framework.  Nevertheless,
because of the changing landscape, there are many research questions
we are only now in position to try to answer: for a theorist,
graph-theoretic questions of three-dimensional enumeration; for a
database designer, more effective search algorithms and ontologies;
for a practical programmer, expansion of the many open-source
cheminformatics projects.  If more chemists think algorithmically, and
more computer scientists think chemically we are much better
positioned taking not only ``simple'' chemicals into account (only one
identifier and isomer possible per molecule), but also their complex
relationships, transformations, and combinatorial challenges.

In 2010, drug company Glaxo made public over 13,000 compounds that
they have investigated as potential cures for
malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
neglected disease in drug research, because the populations who suffer
from it tend not to have enough money to be a stable consumer base for
a new pharmaceutical.)  Perhaps we will now be able to find cures for
some of these neglected diseases, by using the open source, open data
approach to the computer science of chemical discovery.

>>>>

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the 19th
century and flowering in the 1960s with the advent of modern computing
technologies. While many key cheminformatics techniques have been
available in the literature since the 1960's, the bulk of
cheminformatics tools and implementations have been proprietary. In
addition, most data sources of chemical information have been
proprietary (usually with restrictive licensing policies) until
recently. There are many reasons for this, primarily revolving around
issues of profits, competitive intelligence and intellectual property,
which are relevant to the pharmaceutical as well as other chemical
industries. This is in contrast to much of bioinformatics, where the
data and tools have been freely available since the inception of the
field. One could argue that the outcomes from cheminformatics methods
and software have a more direct impact on profits, in terms of
identifying lead like compounds and improving properties of drug
candidates. In contrast, much bioinformatics software is employed in
upstream areas such as target identification.  While certainly
important, this type of research is not as directly related to
possible profits as an actual small molecule.

Furthermore, chemical data (structure, activity information etc.) is
harder to aquire and when done on large scales can be extremely
intensive in terms of time and effort. Thus much of it has been
proprietary. Whereas a lot of bioinformatics data (sequences for
example) are much easier to obtain. While both fields have theoretical
components, it is possible that the free availability of large amounts
of data in bioinformatics drove the development of publically
available tools to handle it. In contrast, the proprietary nature of
chemical data would imply that tools to process and analyse that data
were of interest primarily to the owners. As a result, the incentive
to making cheminformatics software public has been low and since the
users of it were primarily industrial, commercial software was the
norm.

Of course, with the advent of systems biology and its large amounts of
heterogenous data and wide array of (possibly profitable) commercial
applications and, this may change. Indeed, in this field, the
requriement to consider small molecules in the context of larger
biosystems implies that cheminformatics and bioinformatics will both
be required to address the problems that will be tackled.

As we enter the second decade of the 21st century, the cheminformatics
landscape is rapidly shifting.  There now exist free-of-charge
open-source cheminformatics toolkits, applications and open databases
with tens of millions of compounds and containing valuable
experimental bioactivity data. Modern experimental techniques such as
high throughput sequencing have made the generation of large amounts
of structure-activity data much more accessible.  We believe the
prospects for academic and industrial collaboration between chemists
and computer scientists are bright, and we would like to encourage
more computer scientists to participate in cheminformatics research.

There is an admittedly steep learning curve, as significant domain
knowledge is required to attack problems of significance. Indeed, many
issues faced by chemists are often detailed and subtle and do not
admit as clean an abstraction as ``algorithms on a string'', the way
many bioinformatics algorithms can be abstracted into a computer
science framework.  As a result, while one can certainly make
significant contributions to cheminformatics by just considering
structures as graphs, one is limited to somewhat abstract
problems. Keeping in mind that cheminformatics is fundamentally a
practical field that serves to help and advance experimental
chemistry, the key challenges require an understanding of the
underlying chemical systems. One could certainly study for a degree in
chemistry. Alternatively one could collaborate with a chemist or
cheminformatics practitioner.

Nevertheless, because of the increasing availability of tools and
data, the barrier to entry for non-chemists and non-cheminformaticians
is significantly lower. Many questions that benefit from computer
scientists can now be comprehensively taclked: for a theorist,
graph-theoretic questions of three-dimensional enumeration; for a
database designer, more effective search algorithms and ontologies;
for a practical programmer, expansion of the many open-source
cheminformatics projects.  If more chemists think algorithmically, and
more computer scientists think chemically we are much better
positioned taking not only ``simple'' chemicals into account (only one
identifier and isomer possible per molecule), but also their complex
relationships, transformations, and combinatorial challenges.
